The influence of small reservoirs on hydrological drought propagation in space and time

### 2022-10-24 LateX/Feb-3-22-latex-templates/paper_Colombo.tex
new version
started LateX work

\authors{P. Colombo\affil{1,3}, G. G. Ribeiro Neto\affil{2}, P. R. Van Oel\affil{3}, A. C. Costa\affil{4}}
%, G. Mamede\affil{5}

\affiliation{1}{Department of Civil and Environmental Engineering, Politecnico di Milano, Milan, Italy}
\affiliation{2}{Hydrology and Quantitative Water Management, Wageningen University, Wageningen, The Netherlands}
\affiliation{3}{Water Resources Management Group, Wageningen University, Wageningen, The Netherlands}
\affiliation{4}{Institute of Engineering and Sustainable Development, Universidade da Integração Internacional da Lusofonia Afro-Brasileira (Unilab), Redenção, Brazil}
%\affiliation{5}{Institute for Engineering and Sustainable Development, Univ. of International Afro-Brazilian Integration}

@@

\correspondingauthor{Paolo Colombo}{paolo1.colombo@mail.polimi.it}

@@

\begin{keypoints}
\item We evaluated how a dense network of small and unmonitored reservoirs (DRN) impacts drought events in time and space
\item Drought cycle analyses and the downstreamness concept were coupled with hydrological modeling-based DNR scenarios to assess the impacts in time and space, respectively
\item The presence of small reservoirs accelerated the transition towards hydrological droughts by 20\% on average and slowed down drought recovery by 25\%, while shifting upstream water storage capacity distribution in space by 8\%
\end{keypoints}

@@

\begin{abstract}
To increase drought preparedness in semi-arid regions many small and medium reservoirs have been built in recent decades. Together these reservoirs form a Dense Reservoir Network (DRN) and its presence generates numerous challenges for water management. Most of the reservoirs that constitute the network are unmonitored and unregistered, posing questions on their cumulative effects on centralized reservoirs and water distribution at watershed scale. Their influence on hydrological drought propagation is thus largely unexplored. The objective of this study is then to assess the DRN effects on droughts both in time and space. This study utilized a mesoscale semi-distributed hydrological model to reproduce the DRN in a large-scale tropical semiarid watershed (19,530 $km^2$), which presents both a network of large centralized reservoirs and a DRN. To investigate the effects in time and space generated by the network’s presence, the differences between multiple network scenarios were analyzed. Results show that the presence of the DRN accelerates the transition from meteorological to hydrological drought phases by 20\% on average and slows down the recharge in centralized reservoirs by 25\%, leading to a 12\% increase of periods in hydrological drought conditions in a highly centralized basin and 26\% without centralized reservoirs. In space, the DRN shifts upstream the basin’s water storage capacity by 8\%, but when both large and small reservoirs are present the stored volume distribution behavior is not straightforward. The findings confirm the need to consider small reservoirs when addressing drought management policies at regional scale.
\end{abstract}

@@

Droughts are defined as periods of exceptional lack of water that negatively impact human activities or environmental demands. (Van Loon et al., 2016). When drought hits, the most common effect is water shortages, which can lead to food shortages if agriculture is affected. Additional drought impacts include drying rivers and lakes, water use restrictions, dying forests, and wildfires. Combined these can cause famine, diseases, and migration.  Human activities can both aggravate and alleviate hydrological droughts, but the former case has been found more dominant (Van Loon et al., 2022). Identifying droughts’ drivers, both natural and human, is a major scientific challenge (Walker et al., 2022; Zaniolo et al., 2018). Thus, dealing with droughts also requires identifying how humans induce and modify exposure and vulnerability to drought.\\
Articulating the role of drought preparedness in the context of watershed management areas with long term view can be an important practice to mitigate drought events (Gutiérrez et al., 2014). Reservoirs are built in dryland regions both as a drought preparedness measure and as response to the growing water demand (Rabelo et al., 2021). However, reservoir construction without a holistic watershed management approach can in turn generate higher water demand, and at the same time reduces the incentive for adaptive actions on other levels. Therefore, reservoirs can intensify or even induce drought events, since storing water upstream worsened hydrological drought downstream (van Langen et al., 2021). This, in turn, can result in pressure to build more reservoirs, which may further aggravate the problem (Di Baldassarre et al., 2018). Both large publicly managed reservoirs and smaller privately owned reservoirs can play a role in this process.\\
Many places across the world show forms of dense networks of small reservoirs: Australia (Fowler et al., 2015), France (Habets et al., 2014), Ghana (Annor et al., 2009), North-East Brazil (Mamede et al., 2012), South India (Mialhe, 2008), and Syria (Avisse et al., 2017), for example. The influence of small reservoirs on water distribution and droughts is still not fully understood. They are commonly built without regulation nor monitoring in regions where there is a general lack of observational data. This generates uncertainty on their number and capacity, increasing the difficulty of their analysis (Habets et al., 2018). Small reservoirs, when analyzed individually, are not expected to cause major impacts to a hydrological system, since their maximum storage capacity is some orders of magnitude lower than large centralized reservoirs. However, the accumulated effect of a network of small reservoirs can lead to a 30\% increase in the duration of hydrological droughts (Ribeiro Neto et al., 2022). The investigation of the hydrological impact in time and space of a Dense Reservoir Network (DRN) is then highly relevant, both from a community perspective and from a drought and water management one. Moreover, diagnosing droughts also needs the evaluation of the drought management measures in place (Walker et al., 2022).\\
The aim of this research is to evaluate the spatio-temporal influence of a dense network of (small) reservoirs on drought evolution and its impacts on water availability at catchment scale. At the same time, assess the network modeling in a complex basin. We applied a methodology based on the combination of semi-distributed hydrological modeling with a framework for monitoring the evolution of drought events. This methodology was tested for the Banabuiú watershed in the state of Ceara in Brazil.\\
The impact of reservoirs on drought development has been already researched. However, most of these studies focused on large and medium-sized reservoirs (e.g. van Langen et al., 2021). Although some studies have analyzed the effect of small reservoirs on sediment/water dynamics (e.g. Mamede et al, 2018), the hydrological availability (e.g. Krol et al. 2011), evaporation losses, and streamflow impact (e.g. Wisser et al., 2010, Malveira et al.2012), there is a scientific gap related to the cumulative effect of a DRN on drought in a complex hydrological system (Walker et al., 2022). However, direct consideration of the DRN effects on droughts have started to be considered, such as in Ribeiro Neto et al. (2022), where a novel analysis on this topic is proposed and tested in the Riacho do Sangue watershed in North-East Brazil.\\
Integrating DRN's role in a broader water management perspective which does not only consider large and centralized reservoirs may help answer one of the “unsolved problems in hydrology”, the number 22: “What are the synergies and trade-offs between societal goals related to water management?” (Blöschl et al., 2019).

\section{Materials and Methods}
To assess the impacts on drought evolution and magnitude generated by the DRN presence, differences between case study watershed scenarios, including and excluding the DRN, have been analyzed (2.3). Modeling the Banabuiu basin has been selected as the way to represent these scenarios (2.2). The small reservoirs capacities have been estimated to be able to model the network (2.1).
\subsection{Study area}
%Figure 1: Study area framing. North-East Brazil, Ceará and Banabuiú catchment on the left. On the right, centralized reservoirs location and storage capacity, overlayed to DRN’s density in the Banabuiú basin.

%Table 1:  List of centralized reservoirs, with storage capacity and upstream catchment. The table is ordered by decreasing storage capacity.
The semi-arid region of Brazil (North-East Brazil) occupies about 11\% of the Brazilian territory (1,542,000 $km^2$) and contains 30\% of its population (Marengo et al., 2020). It is a representative example of a region with a high concentration of reservoirs with great variation in size and storage capacity. North-East Brazil (NEB) has a highly irregular spatio-temporal precipitation regime, which means the region is frequently affected by intense drought events. Sixteen out of the last 25 years registered rainfall below normal in the region. In 2010 the most severe drought ever recorded started, which ended in 2018 (Marengo et al. 2016, 2018). The combination of the predominance of soils and geology with low water storage capacity produces a dependency on the superficial storage of water (Marengo, Alves, et al., 2017; Marengo, Torres, \& Alves, 2017). Water supply then relies on the reservoirs that regularize discharges. Their storage capacity varies from small reservoirs used on private properties to large reservoirs used for urban supply, industrial demands, and large irrigation areas (also called hydrosystems). A total of 17,083 reservoirs with surface area greater than 5 ha are located in the NEB. This system has a cumulative capacity of 707.36 billion $m^3$, according to the database from Reservoir Monitoring System of the Brazilian National Water Agency (Agência Nacional de Águas e Saneamento Básico) in 2016 (Nascimento and Ribeiro Neto, 2017). A representative example of this situation is Ceará, a state in semi-arid Brazil, where it is estimated the presence of 105, reservoirs with more than 20 meters in length in 148,886 $km^2$ (FUNCEME, 2021). Ceará presents areas with reservoir concentration higher than 7 reservoirs per $km^2$ (Ribeiro Neto et al., 2022), which greatly exceed other high concentration areas such as India (4.2 reservoirs/$km^2$) and Australia (6.1 reservoirs/$km^2$) (Rabelo et. al, 2021).\\
The study area is a watershed located in the state of Ceará. The basin takes the name from its main river, Rio Banabuiú, and covers an area of approximately 19,000 km2. The river network, starting from the main tributary Rio Quixeramobim, delineates the division of the basin into sub-basins and determines the distribution of the big centralized reservoirs, placed along the network branches. The Banabuiú basin counts 19 monitored reservoirs with storage capacity above $1\cdot10^8 m^3$, with the biggest being Arrojado Lisboa at $1.6\cdot10^9 m^3$. These reservoirs, listed in Table 1, are centrally managed and account for $2.79\cdot10^9 m^3$, 93\% of the overall water storage capacity in the basin. At the same time, a cumulative $2.1\cdot10^8 m^3$ storage capacity is retained in approximately 10,000 small reservoirs, which constitute the DRN and are visualized together with centralized reservoirs present in the basin in Figure 1. Despite their small proportion in relation to total capacity (7.1\%), small reservoirs have an important role in the water supply of small rural communities, whose needs are not met by the large hydrosystems, as the water stored in small reservoirs enables local distribution. In the region, most of the small reservoirs aim to meet the water necessities of subsistence farmers only in the short term (6-8 months in the dry season). However, their presence can reduce the potential yield of large, centralized reservoirs, which may determine a lower possibility to store and distribute water during droughts (Krol et al., 2011).

\subsection{DRN capacity estimation}
The capacity of the DRN-component reservoirs is needed to represent the network in the hydrological model.  Neither capacity nor surface area were available. Therefore, a capacity estimation has been obtained starting from the Joint Research Center’s (JRC) Global Surface Water Explorer, which provides a worldwide database of surface water imagery (Pekel et al., 2016). The maximum water extent raster, which maps the maximum extent of water bodies in an area, was connected with the locations of the DRN s(surveyed in FUNCEME et al. 2021), associating each reservoir location to the nearest water body through a nearest neighbor algorithm. In this way, 10,355 small reservoirs had their maximum area associated. Reservoirs without a corresponding water body in JRC’s representation (approximately 7,000) were left out of the final configuration, shown in Figure 1. The causes of them not being represented may lay in their small size, which is difficult to be detected and validated through JRC’s satellite images, or in inaccuracies in the survey operated by FUNCEME et al. (2021), where small topographic depressions could have been erroneously interpreted as reservoirs, such as natural depressions in lowlands and depressions close to roads. The capacity of each reservoir (V) was then obtained from the area (A) through Molle’s equation (Equation 1), where $\alpha$ and K parameters are equal to average values of 2.7 and 1,500 respectively, as used in literature in North-East Brazil (Mamede et al., 2018, 2012; Molle 1994). In this study, a reservoir is considered small when its capacity is lower or equal to 500,000 m3, since this was the highest storage capacity of the unmonitored reservoirs in the basin.
%\begin{linenomath*}
%\begin{equation}
%	V = K \cdot (\frac{A}{ \alfa \cdot K})^\frac{ \alfa }{ \alfa - 1 }
%\end{equation}
%\end{linenomath*}

\subsection{Modeling approach}
\subsubsection{WASA-SED model}
The model selected to model the Banabuiú river basin is WASA-SED (Water Availability in Semi-Arid environments-SEDiments), which has been developed within the joint Spanish-Brazilian-German research project SESAM (Sediment Export from Semi-Arid Catchments: Measurement and Modelling) and utilized to simulate North-East Brazil’s conditions (Güntner, 2002; Güntner et al., 2004; Mueller et al., 2010). The model simulates the runoff and erosion processes at meso-scale, for domains ranging from several hundreds to thousands of square kilometers. It uses a hierarchical top-down disaggregation scheme, dividing each sub-basin of the model into landscape units (based on the Soil and Terrain Digital Database (SOTER) concept (Oldeman and van Engelen, 1993)), represented by multiple terrain components (defined by slope-gradient, length, soil and soil-vegetation components). The hydrological module is fully described in Güntner (2002) and Güntner et al. (2004), implementing for example equations to account for interception losses, evaporation and transpiration (modified Penman-Monteith approach (Shuttleworth and Wallace, 1985)) and for infiltration (Green-Ampt approach (Green and Ampt, 1911)) and other soil and vegetation related processes. The model can simulate both large centralized reservoirs and small diffuse networks of reservoirs (Güntner et al., 2004). The user manual, the link to the source code and other useful sources can be found at \url{https://github.com/TillF/WASA-SED}. Detailed WASA-SED model parameterization for Banabuiú river basin was carried out by Costa et al. (2013), and adopted in this work.

\subsubsection{Small reservoirs representation}
The modeling approach to the reservoirs needs to classify them into strategic and small reservoirs, according to location and size (Güntner et al., 2004). The centralized reservoirs listed in Table 1 are set as strategic and individually parameterized in WASA-SED, also providing their daily regulated discharge time series. Their water balance is calculated explicitly and individually for each reservoir. Due to the uncontrolled and unmonitored nature of small reservoirs composing the DRN, not enough information is available concerning dam building, location, size and water use. They are grouped in 5 storage capacity classes and the water balance is computed for a hypothetical reservoir with mean characteristics, representative of each class and sub-basin. The water storage volume is then given by the product of the water storage of the representative reservoir by the total number of reservoirs from that class within a sub-basin. The generated runoff of the sub-basin is distributed among the reservoir classes through a cascade routing scheme starting from the lowest class to the highest, using a weighting factor computed as a ratio between the runoff contributing area of that reservoir class and the sub-basin area (Mamede et al., 2018). A detailed description of the calculation of water fluxes through the classes of reservoirs can be found in Mamede (2008), while equations and other information about the hydrological and reservoir module can be found in Güntner (2002).

\subsubsection{Calibration and validation}
Calibration was focused on the “scaling factor” parameter, which modifies the soil saturated hydraulic conductivity, which is the most sensitive parameter  (Güntner et al., 2002; Güntner et al., 2021). The calibration was carried out automatically following the sub-basins routing path of the model from upstream to downstream, testing a set of 35 values on each sub-basin (ranging from 0.2 to 7). For each run, five model performance indexes were computed (R2, NSE, KGE, PBIAS and NRMSE) comparing the modeled volume time series for the sub-basin’s centralized reservoir (or the direct downstream one in case the sub-basin did not present one itself) with the observed one. The performance indexes were selected based on recent usage in hydrological modeling (Marahatta et al., 2021; Knoben et al., 2019; Uniyal et al., 2019). The parameter that would produce the best performance was selected. The model was calibrated with 70\% of the series, running it from 1980 to 2006, it was then validated on the remaining 30\% of the series, from 2007 to 2018. Wet, normal and dry years were well represented in both the selected periods (calibration and validation) of the gauges. Therefore, the high inter-annual streamflow variability (CV greater than 1) was taken into account for both calibration and validation periods.

>>>>

\authors{P. Colombo\affil{1,3}, G. G. Ribeiro Neto\affil{2}, P. R. Van Oel\affil{3}, A. C. Costa\affil{4}, G. Mamede\affil{4}}

\affiliation{1}{Department of Civil and Environmental Engineering, Politecnico di Milano, Milan, Italy}
\affiliation{2}{Hydrology and Quantitative Water Management, Wageningen University, Wageningen, The Netherlands}
\affiliation{3}{Water Resources Management Group, Wageningen University, Wageningen, The Netherlands}
\affiliation{4}{Institute of Engineering and Sustainable Development, Universidade da Integração Internacional da Lusofonia Afro-Brasileira (Unilab), Redenção, Brazil}

@@

\correspondingauthor{Germano G. Ribeiro Neto}{germano.gondimribeironeto@wur.nl}

@@

\begin{keypoints}
\item We evaluated how a dense network of small and unmonitored reservoirs (DRN) impacts drought events in time and space
\item Drought cycle analyses and the downstreamness concept were coupled with hydrological modeling-based DRN scenarios to assess the impacts in time and space, respectively
\item The presence of small reservoirs accelerated the transition towards hydrological droughts by 20\% on average and slowed down drought recovery by 25\%, while shifting upstream water storage capacity distribution in space by 8\%
\end{keypoints}

@@

\begin{abstract}
To increase drought preparedness in semi-arid regions many small and medium reservoirs have been built in recent decades. Together these reservoirs form a Dense Reservoir Network (DRN) and its presence generates numerous challenges for water management. Most of the reservoirs that constitute the network are unmonitored and unregistered, posing questions on their cumulative effects on strategic reservoirs and water distribution at watershed scale. Their influence on hydrological drought propagation is thus largely unexplored. The objective of this study is then to assess the DRN effects on droughts both in time and space. This study utilized a mesoscale semi-distributed hydrological model to reproduce the DRN in a large-scale tropical semiarid watershed (19,530 km2), which presents both a network of large strategic reservoirs and a DRN. To investigate the effects in time and space generated by the network’s presence, the differences between multiple network scenarios were analyzed. Results show that the presence of the DRN accelerates the transition from meteorological to hydrological drought phases by 20\% on average and slows down the recharge in strategic reservoirs by 25\%, leading to a 12\% increase of periods in hydrological drought conditions in a highly strategic basin and 26\% without strategic reservoirs. In space, the DRN shifts upstream the basin’s water storage capacity by 8\%, but when both large and small reservoirs are present the stored volume distribution behavior is not straightforward. The findings confirm the need to consider small reservoirs when addressing drought management policies at regional scale.
\end{abstract}

@@

Droughts are defined as periods of exceptional lack of water that negatively impact human activities or environmental demands. (Van Loon et al., 2016). When drought hits, the most common effect is water shortages, which can lead to food shortages if agriculture is affected. Additional drought impacts include drying rivers and lakes, water use restrictions, reduced electricity production, dying forests, and wildfires. Combined these can cause famine, diseases, and migration.  Human activities can both aggravate and alleviate hydrological droughts, but the former case has been found more dominant (Van Loon et al., 2022). Identifying droughts’ drivers, both natural and human, is a major scientific challenge (Walker et al., 2022; Zaniolo et al., 2018). Thus, dealing with droughts also requires identifying how humans induce and modify exposure and vulnerability to drought.\\
Articulating the role of drought preparedness in the context of watershed management areas with a long term view can be an important practice to mitigate drought events (Gutiérrez et al., 2014). Reservoirs are built in dryland regions both as a drought preparedness measure and as response to the growing water demand (Rabelo et al., 2021). However, reservoir construction without a holistic watershed management approach can in turn generate higher water demand, and at the same time reduces the incentive for adaptive actions on other levels. Therefore, reservoirs can intensify or even induce drought events, since storing water upstream worsened hydrological drought downstream (Van Langen et al., 2021). This, in turn, can result in pressure to build more reservoirs, which may further aggravate the problem (Di Baldassarre et al., 2018). Both large publicly managed reservoirs and smaller privately owned reservoirs can play a role in this process.\\
Many places across the world show forms of dense networks of small reservoirs: Australia (Fowler et al., 2015), France (Habets et al., 2014), Ghana (Annor et al., 2009), North-East Brazil (Mamede et al., 2012), South India (Mialhe, 2008), and Syria (Avisse et al., 2017), for example. The influence of small reservoirs on water distribution and droughts is still not fully understood. They are commonly built without regulation nor monitoring in regions where there is a general lack of observational data. This generates uncertainty on their number and capacity, increasing the difficulty of their analysis (Habets et al., 2018). Small reservoirs, when analyzed individually, are not expected to cause major impacts to a hydrological system, since their maximum storage capacity is some orders of magnitude lower than strategic reservoirs (medium and large-sized reservoirs located on main rivers at the sub-basin’s outlet). However, the accumulated effect of a network of small reservoirs can lead to a 30\% increase in the duration of hydrological droughts (Ribeiro Neto et al., 2022). The investigation of the hydrological impact in time and space of a Dense Reservoir Network (DRN) is then highly relevant, both from a community perspective and from a drought and water management one. Moreover, diagnosing droughts also needs the evaluation of the drought management measures in place (Walker et al., 2022).\\
The impact of reservoirs on drought development has been already researched. However, most of these studies focused on large and medium-sized reservoirs (e.g. van Langen et al., 2021). Although some studies have analyzed the effect of small reservoirs on sediment/water dynamics (e.g. Mamede et al, 2018), the hydrological availability (e.g. Krol et al. 2011), evaporation losses, and streamflow impact (e.g. Wisser et al., 2010, Malveira et al.2012), there is a scientific gap related to the cumulative effect of a DRN on drought in a complex hydrological system (Walker et al., 2022). However, direct consideration of the DRN effects on droughts have started to be considered, such as in Ribeiro Neto et al. (2022), where a novel analysis on this topic is proposed and tested in the Riacho do Sangue watershed in North-East Brazil.Integrating DRN's role in a broader water management perspective which does not only consider strategic centralized reservoirs may help answer one of the “unsolved problems in hydrology”, the number 22: “What are the synergies and trade-offs between societal goals related to water management?” (Blöschl et al., 2019).\\
The aim of this research is to evaluate the spatio-temporal influence of a dense network of (small) reservoirs on drought evolution and its impacts on water availability at catchment scale. At the same time, assess the network modeling in a complex basin. We applied a methodology based on the combination of semi-distributed hydrological modeling with a framework for monitoring the evolution of drought events. This methodology was tested for the Banabuiú watershed in the state of Ceara in Brazil.

\section{Materials and Methods}
To assess the impacts on drought evolution and magnitude generated by the DRN presence, differences between case study watershed scenarios, including and excluding the DRN, have been analyzed (2.3). Modeling the Banabuiu basin has been selected as the way to represent these scenarios (2.2). The small reservoirs capacities have been estimated to be able to model the network (2.1).
\subsection{Study area}
%Figure 1: Study area framing. North-East Brazil, Ceará and Banabuiú catchment on the left. On the right, centralized reservoirs location and storage capacity, overlayed to DRN’s density in the Banabuiú basin.
The semi-arid region of Brazil (North-East Brazil) occupies about 11\% of the Brazilian territory (1,542,000 $km^2$) and contains 30\% of its population (Marengo et al., 2020). It is a representative example of a region with a high concentration of reservoirs with great variation in size and storage capacity. North-East Brazil (NEB) has a highly irregular spatio-temporal precipitation regime, which means the region is frequently affected by intense drought events. Sixteen out of the last 25 years registered rainfall below normal in the region. In 2010 the most severe drought ever recorded started, which ended in 2018 (Marengo et al. 2016, 2018). The combination of the predominance of soils and geology with low water storage capacity produces a dependency on the superficial storage of water (Marengo, Alves, et al., 2017; Marengo, Torres, & Alves, 2017). Water supply then relies on the reservoirs that regularize discharges. Their storage capacity varies from small reservoirs used on private properties to large reservoirs used for urban supply, industrial demands, and large irrigation areas (also called hydrosystems). A total of 17,083 reservoirs with surface area greater than 5 ha are located in the NEB. This system has a cumulative capacity of 707.36 billion $m^3$, according to the database from Reservoir Monitoring System of the Brazilian National Water Agency (Agência Nacional de Águas e Saneamento Básico) in 2016 (Nascimento and Ribeiro Neto, 2017). A representative example of this situation is Ceará, a 148,886 km² wide state in semi-arid Brazil, where it is estimated the presence of 105,813 reservoirs with more than 20 meters in length (FUNCEME, 2021). Ceará presents areas with reservoir concentration higher than 7 reservoirs per $km^2$ (Ribeiro Neto et al., 2022), which greatly exceed other high concentration areas such as India (4.2 reservoirs/$km^2$) and Australia (6.1 reservoirs/$km^2$) (Rabelo et. al, 2021).\\
The study area is the Banabuiu river watershed located in the state of Ceará which covers an area of approximately 19,000 $km^2$. The river network, starting from the main tributary Rio Quixeramobim, delineates the division of the basin into sub-basins and determines the distribution of the strategic reservoirs, placed along the network branches. The Banabuiú basin counts 19 monitored reservoirs with storage capacity above 1·$Hm^3$ ($10^6 m^3$) , with the biggest being Arrojado Lisboa at 1600 $Hm^3$. These reservoirs, shown in Figure 1, are continuously monitored and managed by the State water agency (Water Resources Company, COGERH) and account for 2790 $Hm^3$, 93\% of the overall water storage capacity in the basin. At the same time, a cumulative 210 $Hm^3$ storage capacity is retained in approximately 10,000 small reservoirs, which constitute the DRN and are visualized together with strategic reservoirs present in the basin in Figure 1. Despite their small proportion in relation to total capacity (7.1\%), small reservoirs have an important role in the water supply of small rural communities, whose needs are not met by the large hydrosystems, as the water stored in small reservoirs enables local distribution. In the region, most of the small reservoirs aim to meet the water necessities of subsistence farmers only in the short term (6-8 months in the dry season). The small reservoirs also serve as sediment detention basins, retaining a considerable amount of sediment generated within the catchment and extending the life-time of larger ones located downstream (MAMEDE et al. 2018). However, their presence can reduce the potential yield of strategic reservoirs which may determine a lower possibility to store and distribute water during droughts (Krol et al., 2011).

\subsection{DRN capacity estimation}
%Table 1. Repartition of the reservoirs composing the Dense Reservoir Network in the Banabuiu basin.
The capacity of the DRN-component reservoirs is needed to represent the network in the hydrological model.  Neither capacity nor surface area were available. Therefore, a capacity estimation has been obtained starting from the Joint Research Center’s (JRC) Global Surface Water Explorer, which provides a worldwide database of surface water imagery (Pekel et al., 2016). The maximum water extent raster, which maps the maximum extent of water bodies in an area, was connected with the locations of the DRN (surveyed in FUNCEME et al. 2021), associating each reservoir location to the nearest water body through a nearest neighbor algorithm. In this way, 10,355 small reservoirs had their maximum area associated. Reservoirs without a corresponding water body in JRC’s representation (approximately 7,000) were left out of the final configuration, represented in Figure 1 and summarized in Table 1. The causes of them not being represented may lay in their small size, which is difficult to be detected and validated through JRC’s satellite images, or in inaccuracies in the survey operated by FUNCEME et al. (2021), where small topographic depressions could have been erroneously interpreted as reservoirs, such as natural depressions in lowlands and depressions close to roads. The capacity of each reservoir (V) was then obtained from the area (A) through Molle’s equation (Equation 1), where α and K parameters are equal to average values of 2.7 and 1,500 respectively, as used in literature in North-East Brazil (Mamede et al., 2018, 2012; Molle 1994). In this study, a reservoir is considered small when its capacity is lower or equal to 500,000 $m^3$ (0.5 $Hm^3$), since this was the highest storage capacity of the unmonitored reservoirs in the basin.
%\begin{linenomath*}
%\begin{equation}
%	V = K \cdot (\frac{A}{ \alfa \cdot K})^\frac{ \alfa }{ \alfa - 1 }
%\end{equation}
%\end{linenomath*}

\subsection{Modeling approach}
\subsubsection{WASA-SED model}
The model selected to model the Banabuiú river basin is WASA-SED (Water Availability in Semi-Arid environments-SEDiments), which has been developed within the joint Spanish-Brazilian-German research project SESAM (Sediment Export from Semi-Arid Catchments: Measurement and Modelling) and utilized to simulate North-East Brazil’s conditions (Güntner, 2002; Güntner et al., 2004; Mueller et al., 2010). The model simulates the runoff and erosion processes at meso-scale, for domains ranging from several hundreds to thousands of square kilometers. It uses a hierarchical top-down disaggregation scheme, dividing each sub-basin of the model into landscape units (based on the Soil and Terrain Digital Database (SOTER) concept (Oldeman and van Engelen, 1993)), represented by multiple terrain components (defined by slope-gradient, length, soil and soil-vegetation components). The hydrological module is fully described in Güntner (2002) and Güntner et al. (2004), implementing for example equations to account for interception losses, evaporation and transpiration (modified Penman-Monteith approach (Shuttleworth and Wallace, 1985)) and for infiltration (Green-Ampt approach (Green and Ampt, 1911)) and other soil and vegetation related processes. The model can simulate both medium and large  centralized strategic reservoirs and small diffuse networks of reservoirs (Güntner et al., 2004). The user manual, the link to the source code and other useful sources can be found at \url{github.com/TillF/WASA-SED}. Detailed WASA-SED model parameterization for Banabuiú river basin was carried out by Costa et al. (2013), and adopted in this work.

\subsubsection{Small reservoirs representation}
The modeling approach to the reservoirs needs to classify them into strategic and small reservoirs, according to location and size (Güntner et al., 2004). The strategic reservoirs listed in Table 1 are individually parameterized in WASA-SED, also providing their daily regulated discharge time series. Their water balance is calculated explicitly and individually for each reservoir. Due to the uncontrolled and unmonitored nature of small reservoirs composing the DRN, not enough information is available concerning dam building, location, size and water use. They are grouped into 5 storage capacity classes and the water balance is computed for a hypothetical reservoir with mean characteristics, representative of each class and sub-basin. The water storage volume is then given by the product of the water storage of the representative reservoir by the total number of reservoirs from that class within a sub-basin. The generated runoff of the sub-basin is distributed among the reservoir classes through a cascade routing scheme starting from the lowest class to the highest, using a weighting factor computed as a ratio between the runoff contributing area of that reservoir class and the sub-basin area (Mamede et al., 2018). A detailed description of the calculation of water fluxes through the classes of reservoirs can be found in Mamede (2008), while equations and other information about the hydrological and reservoir module can be found in Güntner (2002).

\subsubsection{Calibration and validation}
Calibration was focused on the “scaling factor” parameter, which modifies the soil saturated hydraulic conductivity, which is the most sensitive parameter  (Güntner et al., 2002; Güntner et al., 2021). The calibration was carried out automatically following the sub-basins routing path of the model from upstream to downstream, testing a set of 35 values on each sub-basin (ranging from 0.2 to 7). For each run, five model performance indexes were computed (R2, NSE, KGE, PBIAS and NRMSE) comparing the modeled volume time series for the sub-basin’s strategic reservoir (or the direct downstream one in case the sub-basin did not present one itself) with the observed one. The performance indexes were selected based on recent usage in hydrological modeling (Marahatta et al., 2021; Knoben et al., 2019; Uniyal et al., 2019). The parameter that would produce the best performance was selected. The model was calibrated with 70\% of the series, running it from 1980 to 2006, it was then validated on the remaining 30\% of the series, from 2007 to 2018. Wet, normal and dry years were well represented in both the selected periods (calibration and validation) of the gauges. Therefore, the high inter-annual streamflow variability (CV greater than 1) was taken into account for both calibration and validation periods.
